---
title: "Logistic regression and Dose Response Data"
output: pdf_document
Author: Joel Steele
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(comment='')
```

Many thanks to Thaddeus Tarpey at Wright University
 Check out his cite for this and more   
 http://www.wright.edu/~thaddeus.tarpey/   

 These data are a reproduction of data from
 C.I. Bliss (1935). The calculation of the dosage-mortality curve.
 _Annals of Applied Biology_, vol 22, Issue 1, 134-167.
      
   
### Data
Beetles were exposed to carbon disulphide at varying concentrations for $5$ hours.   
   
* dose = mf/L concentration of $CS_2$
* nexp = number of beetles exposed
* ndied = number of beetles killed
* prop = proportion of dead to exposed beetles 
   

```{r}
exp.dat = matrix(c(49.1,53,56.9,60.8,64.8,68.7,72.6,76.5,
                   59,60,62,56,63,59,62,60,
                   6,13,18,28,52,53,61,60,
                   .102,.217,.29,.5,.825,.898,.984,1),ncol=4)
colnames(exp.dat) = c('dose','nexp','ndied','prop')
exp.dat = as.data.frame(exp.dat)
# compute how many lived
exp.dat$nalive = exp.dat$nexp - exp.dat$ndied
```

### Logistic Model
Run a logistic regression of the proportion of dead to living beetles as a function of the dose of $CS_2$ gas.

```{r}
summary(glm(cbind(ndied,nalive) ~ dose, 
            family=binomial,
            data=exp.dat)->exp.glm)
```

  
we may be interested in finding the concentration of $CS_2$ gas  that is lethal 50% of the time, the $LD_{50}$  below is the function for this computation.
   
```{r}
dose4prob = function(b0,b1,prob){
  d = (-b0+log(-prob/(prob-1)))/b1
  return(d)
} 
dose4prob(coef(exp.glm)[[1]],coef(exp.glm)[[2]],.5)
```
# Note that if we have a function with multiple predictors we
 can solve for each variable using something similar.
 For example if
 $$y \sim b_0 + b_1(x_1) + b_2(x_2) + b_3(x_3)$$   
 is the model. Then to find a specific value for one of the predictors ($x_1, x_2, x_3$) that corresponds to a desired probability ($y$).

* $x_1 = (-b_0-b_2-b_3+log \left(\frac{-y}{(y-1)}\right))/b1$
* $x_2 = (-b_0-b_1-b_3+log \left(\frac{-y}{(y-1)}\right))/b2$
* $x_3 = (-b_0-b_1-b_2+log \left(\frac{-y}{(y-1)}\right))/b3$
    
```{r}
# what is the range of doses
range(exp.dat$dose)
# let's create our own range to predict with
drange = seq(30,90,length=100)
# predict probability of beetle death based on the model
exp.pred = predict(exp.glm, newdata=data.frame(dose=drange))
# Now we plot
# first the predicted values
plot(drange, 
     exp(exp.pred)/(1+exp(exp.pred)),
     type='l',
     xlab='dose',
     ylab='probability')
# add in the observed points
points(prop ~ dose, data = exp.dat, col='red', pch=19)
# compute our LD values
ld50 = dose4prob(coef(exp.glm)[[1]],coef(exp.glm)[[2]],.50)
ld25 = dose4prob(coef(exp.glm)[[1]],coef(exp.glm)[[2]],.25)
# let's add these to our graph for visualization purposes
abline(h=.5,lty='dashed',col='red')
abline(h=.25,lty='dotted',col='blue')
legend('topleft',
       c(expression(LD[50]),
         expression(LD[25])),
       col=c('red','blue'),lty=c('dashed','dotted'),
       inset=.01)
text(ld50,.52,sprintf("%.3f",ld50),pos=c(2),col='red')
text(ld25,.27,sprintf("%.3f",ld25),pos=c(2),col='blue')
```
